Why do people punish? Evidence for a range of strategic concerns

### 2024-05-22 manuscript.tex
submit manuscript

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  man,floatsintext]{apa6}
\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\ifLuaTeX
\usepackage[bidi=basic]{babel}
\else
\usepackage[bidi=default]{babel}
\fi
\babelprovide[main,import]{english}
% get rid of language-specific shorthands (see #6817):
\let\LanguageShortHands\languageshorthands
\def\languageshorthands#1{}
% Manuscript styling
\usepackage{upgreek}
\captionsetup{font=singlespacing,justification=justified}

@@

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi\\: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\keywords{punishment; cooperation; economic games\newline\indent Word count: 5657 words}
\usepackage{lineno}

\linenumbers
\usepackage{csquotes}
\usepackage{array}
\usepackage{caption}
\usepackage{graphicx}
\usepackage{siunitx}
\usepackage[normalem]{ulem}
\usepackage{colortbl}
\usepackage{multirow}
\usepackage{hhline}
\usepackage{calc}
\usepackage{tabularx}
\usepackage{threeparttable}
\usepackage{wrapfig}
\usepackage{adjustbox}
\usepackage{hyperref}
\usepackage{setspace}
\usepackage{amsmath}
\usepackage{xcolor}
\usepackage{makecell}
\raggedbottom
\AtBeginEnvironment{tabular}{\singlespacing}
\AtBeginEnvironment{lltable}{\singlespacing}
\AtBeginEnvironment{tablenotes}{\doublespacing}
\captionsetup[table]{font={stretch=1,small}}
\captionsetup[figure]{font={stretch=1,small}}
\nolinenumbers
\note{\raggedright * Correspondence concerning this article should be addressed to Scott Claessens, Level 2, Building 302, 23 Symonds Street, Auckland, New Zealand. \text{E-mail:} \href{scott.claessens@gmail.com}{\nolinkurl{scott.claessens@gmail.com}} \par This working paper has not yet been peer-reviewed.}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Why do people punish? Evidence for a range of strategic concerns},
  pdfauthor={Scott Claessens*1, Quentin D. Atkinson1, \& Nichola Raihani1,2},
  pdflang={en-EN},
  pdfkeywords={punishment; cooperation; economic games},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

@@

\nolinenumbers

@@

Humans cooperate on a scale that is unparalleled in the animal kingdom. One
mechanism thought to sustain this level of cooperation is costly punishment,
whereby individuals harm others at a personal cost\textsuperscript{1},
ostensibly encouraging cooperative behaviour from the target (or bystanders\textsuperscript{2--4}) in the future. Punishment offers a
route to maintaining or increasing cooperation by changing the payoff structure
of social interactions such that it no longer pays to cheat or exploit social
partners\textsuperscript{1,5}. Yet, despite its theoretical
importance, the question of why people choose to punish others is still hotly
contested\textsuperscript{6}. In this study, we use a battery of economic games to
disentangle the different motives underpinning punishment and explore how these
motives vary across individuals.

In humans, many studies of punishment have been carried out in laboratory
settings using economic games\textsuperscript{6--15}. In these games, participants are usually given a sum of money
that they can use to invest in collective action or to help others.
Alternatively, participants can `cheat' by keeping the money for themselves or
by exploiting the contributions of others. Punishment is introduced into such
games by giving participants the option to pay a small `fee' to impose a greater
`fine' on their co-players. Several lines of experimental evidence indicate that
people use this punishment option\textsuperscript{13}, that they enjoy
punishing\textsuperscript{16}, and that they frequently, though not
always\textsuperscript{17}, punish cheating or exploitative
co-players\textsuperscript{11,12}.

Evidence from these experiments suggests that the threat of costly punishment
plays an important role in promoting human cooperation. People tend to cooperate
more in games where punishment is possible compared to those where it is not\textsuperscript{6--8}. The effect that the threat of
punishment has on cooperation is also evident in the higher contributions
typically observed in the Ultimatum Game (where punishment is possible) compared
to the structurally-similar Dictator Game (where it is not)\textsuperscript{18}. This
typical cooperation-enhancing effect of punishment has also been observed across
societies\textsuperscript{8}, leading some to suggest that costly punishment has
played a key role in the cultural evolution of cooperation in humans\textsuperscript{19--22}.

Nevertheless, it remains unclear whether individuals playing economic games use
punishment as a behaviour-change tool to enforce cooperation or as a means to
achieve other ends. Some have argued that punishment is primarily used to shape
future behaviour, either to deter personal harm\textsuperscript{3,10,23} or to uphold normative standards of cooperative behaviour\textsuperscript{20,21,24--27}.
But while the \emph{threat} of punishment can have a cooperation-enhancing effect,
the \emph{enactment} of this punishment does not consistently deter targets from
cheating in the future\textsuperscript{6}. This calls into question whether
punishment primarily operates as a behaviour-change tool or whether it is used
to achieve other goals.

Beyond behaviour shaping concerns, there are a host of other reasons why people
may want to punish in economic games. Punishers might be motivated by a desire
for retribution rather than deterrence, punishing in proportion to the amount
of harm that was personally caused\textsuperscript{28}. Punishment might be driven
by concerns about relative payoffs, such as disadvantageous inequity aversion
(i.e., avoiding having less than others\textsuperscript{6,29}) and/or
general egalitarian preferences (i.e., wanting all participants to receive the
same payoffs\textsuperscript{30}). Such concerns about relative payoffs may be activated
when participants earn less than cheats in economic games or when there are
income disparities in these settings. People might also use punishment for
competitive purposes, seeking advantageous inequity for themselves (i.e., having
more than others) and/or improving their relative position\textsuperscript{6}.

Common economic game designs have been unable to tease apart these different
motives for punishment because participants who interact with cheats in these
games experience both losses \emph{and} lower relative payoffs. The typical 1:3
fee-fine ratio of punishment in economic games compounds this issue. With
this setup, people can simultaneously use punishment to reciprocate losses, to
deter others from cheating, and to reduce or reverse disparities in payoffs
between themselves and targets. To add to this complexity, it is evident that
people use punishment in seemingly disparate ways: punishing when no behaviour
change is possible, such as in one-shot games\textsuperscript{13,29,31,32}, on the very last round of repeated games\textsuperscript{33}, or in games where the target never learns about the
punishment\textsuperscript{34}; punishing those who did not cheat or who
over-contributed to collective action (antisocial punishment\textsuperscript{17,35}); punishing in scenarios where they were not
personally harmed (third-party punishment\textsuperscript{36}); and punishing in
scenarios where disparities in payoffs did not arise from participants'
actions\textsuperscript{30,37,38}.

The general conclusion from this research is that there is no one unifying
function of costly punishment in humans. Instead, punishment should be thought
of as a flexible behavioural tool that serves a variety of functions that are
not mutually exclusive\textsuperscript{6}. Due to its multipurpose nature, we should
therefore expect variation in punishment strategies in the population, much like
the observed variation in social learning strategies\textsuperscript{39}. Some
individuals may use punishment as a behaviour shaping tool, for example, while
others may use it to reduce or reverse payoff differentials.

This insight raises several underexplored questions. First, which punishment
strategies are more frequent in human populations? Second, what traits predict
adherence to a particular strategy? Previous work has reported that personality
is related to cooperative behaviour\textsuperscript{40} and demographics, political
ideology, and religiosity are related to punitive behaviour\textsuperscript{41}, but
no research has related these variables to specific punishment strategies.
Third, do people have insight into their own punishment strategy? Previous work
has argued that people are often unaware of the underlying function of their
punitive behaviour, yet they feel compelled to enact it anyway\textsuperscript{28,42}.

Here, we aim to delineate nine possible punishment strategies by asking
whether people punish in a manner consistent with a specific strategy and, if
so, what other characteristics (personality, social preferences, political
orientation, religious views) predict the use of different punishment strategies.
Table \ref{tab:tableStrategies} summarises the potential functions for costly
punishment in the economic games that we considered, and the behavioural
strategies they predict. Note that Table \ref{tab:tableStrategies} is not
an exhaustive list of all possible punishment strategies: we do not include
reputational functions of punishment in this table, such as signalling
trustworthiness\textsuperscript{4,43--46},
because our focus is on punishment strategies in anonymous economic games without
reputational incentives (but see ref\textsuperscript{47}).

Building on previous designs\textsuperscript{29,31,48,49}, we employ a suite of one-shot economic games where individuals
are given the opportunity to punish targets at a personal cost (Figure
\ref{fig:plotGames}). In each game, targets either steal from another
individual or do nothing. Representative samples of participants from the United
Kingdom (\emph{n} = 1014) and the
United States (\emph{n} = 996)
completed all six games on the online platform Prolific. We carefully designed
the suite of games to tease apart the proposed punishment strategies in Table
\ref{tab:tableStrategies}, such that each strategy predicts a different pattern
of behaviour across all the games (see Methods for more detail about the six
games). We use the resulting behavioural patterns to discern which punishment
strategy participants are employing. We then combine these behavioural patterns
with data on demographics, personality, social preferences, political ideology,
religiosity, and self-reported strategy usage.

@@

Using a suite of economic games measuring punishment in different situations,
we have shown that punishment does not serve just one function, but instead is
a flexible tool that can be and is used for different purposes\textsuperscript{6}.
While some use punishment to enforce norms of cooperation, others use it to
reduce or even create inequality between individuals. We found that people's
punishment strategy can, to some extent, be predicted by individual differences
in personality, social preferences, and political and religious views. Moreover,
contrary to the view that people are often unable to articulate the reasons for
their punitive behaviour\textsuperscript{28}, people seem to have some degree of
insight into the strategy they are using. Despite small differences, these
general patterns replicated in samples from both the United Kingdom and the
United States, providing further confidence in the results.

Among the punitive strategies, the most common were particularly sensitive to
inequality in payoffs, either from a self-referential perspective (i.e., avoid
disadvantageous inequity) or more generally (i.e., egalitarian). This is in
line with previous studies which have highlighted inequity aversion as an
important motivation for punishment in economic games\textsuperscript{29--31,37,48,50}.
Personality, social preferences, and political and religious views were related
to these strategies. Traits associated with other-regarding concern, such as SVO
and honesty-humility, predicted following the egalitarian strategy. By contrast,
religious and conservative individuals, including individuals higher in SDO and
RWA, were less likely to follow the egalitarian strategy and more likely to
follow the avoid disadvantageous inequity strategy, especially in the United
States. Authoritarian participants in the United States were also more likely to
actively seek advantageous inequity with their punishment. These findings
provide insight into the motives that may have driven higher levels of peer
punishment among conservatives in previous work\textsuperscript{41}.

Behaviour shaping strategies, such as deterrence and norm-enforcement, were less
common than strategies sensitive to inequality in our set of games. This was
reflected both in participants' elicited punishment behaviour (Figure
\ref{fig:plotModel1b}) and in their self-reports of their own strategy
(Supplementary Figures \ref{fig:plotSliders1} and \ref{fig:plotSliders2}).
Although our design did not explicitly allow for behaviour shaping as the
interactions were all one-shot, we did manipulate whether the target's stealing
behaviour was intentional or not (Game C), an approach which has been used in
previous vignette studies to identify behaviour shaping motives\textsuperscript{28}.
The lower prevalence of behaviour shaping strategies in our study is consistent
with prior work showing that punishment often continues to be used when
behaviour shaping is impossible, such as when the target will never find out
that they have been punished\textsuperscript{34} or on the last round of repeated
interactions\textsuperscript{33}. We found that demographic, personality, political,
and religious variables tended to be unrelated to behaviour shaping strategies,
though this could reflect low power to detect such associations given the low
prevalence of these strategies in our sample. We also found that participants
accurately reported using the norm-enforcing strategy, but not the deterrent
strategy. This finding is in line with previous research showing that people
struggle to accurately report the deterrent motivations for their punitive
behaviour\textsuperscript{28}.

We defined antisocial punishment as occurring when a participant exclusively
punished those who did not steal. There were no participants in our sample who
followed this strategy, although some players did punish non-stealing
co-players. Harming non-stealing individuals was also consistent with the
competitive strategies, which did appear in our sample albeit at low frequencies.
The fact that no participants in our sample punished non-stealing across all
games suggests that the antisocial punishment that has been observed in other
studies\textsuperscript{17,51} is not aimed at harming cooperators
specifically, as has been previously suggested\textsuperscript{17}. Instead,
antisocial punishment is more likely to be motivated by improving one's
relative position, which is in line with work showing that antisocial punishment
disappears when relative payoffs cannot be changed (e.g., when punishment is
only available with a 1:1 fee-fine ratio\textsuperscript{35}).

The fact that people use punishment for many different reasons poses problems
for the way that punishment is operationalised in classic behavioural economic
game studies. In these studies, a common assumption is that participants punish
to change the behaviour of cheats\textsuperscript{11,12}. But in reality, people
may be choosing the punishment option to achieve a variety of different goals.
This has implications for how people respond to being targeted by punishers in
these games. Targets of punishment in these studies may know that punishment
reflects different motives and can respond accordingly. For example, if targets
interpret punishment as serving a competitive motive, it may elicit retaliation
rather than encourage cooperation\textsuperscript{6,10,52}.
As punishers' motives must be inferred (and such inferences likely depend on
character traits of the target, as well as the context in which punishment
occurs), there is likely to be some variation and error in attributing motives
to punishers. To the extent that inferred motives affect target responses, this
might help to explain the mixed findings in the field as to whether punishment
actually motivates cheating targets to cooperate in the future\textsuperscript{6}.

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  man, donotrepeattitle,floatsintext]{apa6}
\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\ifLuaTeX
\usepackage[bidi=basic]{babel}
\else
\usepackage[bidi=default]{babel}
\fi
\babelprovide[main,import]{english}
% get rid of language-specific shorthands (see #6817):
\let\LanguageShortHands\languageshorthands
\def\languageshorthands#1{}
% Manuscript styling
\usepackage{upgreek}
\captionsetup{font=singlespacing,justification=justified}

@@

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi\\: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\usepackage{lineno}

\linenumbers
\usepackage{csquotes}
\usepackage{array}
\usepackage{caption}
\usepackage{graphicx}
\usepackage{siunitx}
\usepackage[normalem]{ulem}
\usepackage{colortbl}
\usepackage{multirow}
\usepackage{hhline}
\usepackage{calc}
\usepackage{tabularx}
\usepackage{threeparttable}
\usepackage{wrapfig}
\usepackage{adjustbox}
\usepackage{hyperref}
\usepackage{setspace}
\usepackage{amsmath}
\usepackage{xcolor}
\usepackage{makecell}
\raggedbottom
\AtBeginEnvironment{tabular}{\singlespacing}
\AtBeginEnvironment{lltable}{\singlespacing}
\AtBeginEnvironment{tablenotes}{\doublespacing}
\captionsetup[table]{font={stretch=1,small}}
\captionsetup[figure]{font={stretch=1,small}}
\nolinenumbers
\note{\raggedright * Correspondence concerning this article should be addressed to Scott Claessens, Level 2, Building 302, 23 Symonds Street, Auckland, New Zealand. \text{E-mail:} \href{scott.claessens@gmail.com}{\nolinkurl{scott.claessens@gmail.com}} \par This manuscript has not yet been peer-reviewed. \par This manuscript was published as a pre-print on PsyArXiv under a CC BY 4.0 \text{license:} \url{https://doi.org/10.31234/osf.io/ys6rm} \par \text{Classification:} Social Sciences; Psychological and Cognitive Sciences \par \text{Keywords:} punishment; cooperation; economic games}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Why do people punish? Evidence for a range of strategic concerns},
  pdfauthor={Scott Claessens*1, Quentin D. Atkinson1, \& Nichola Raihani1,2},
  pdflang={en-EN},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

@@

\hypertarget{significance-statement}{%
\section{Significance Statement}\label{significance-statement}}

Research has shown that humans are willing to punish others at a personal cost,
but the motives for this behaviour remain unclear. Using a suite of economic
games, we show that punitive behaviour is consistent with a variety of strategic
concerns. While most people choose not to punish in our games, people who do
punish tend to be motivated by egalitarianism rather than deterrence or
norm enforcement. We also find that different punishment strategies are linked
with different character traits, including personality, prosocial preferences,
and political and religious views. This research highlights the multifaceted
nature of human punishment.

\newpage
\linenumbers

@@

Humans cooperate on a scale that is unparalleled in the animal kingdom. One
mechanism thought to sustain this level of cooperation is costly punishment,
whereby individuals harm others at a personal cost (1),
ostensibly encouraging cooperative behaviour from the target, or bystanders
(2--4), in the future. Punishment offers a
route to maintaining or increasing cooperation by changing the payoff structure
of social interactions such that it no longer pays to cheat or exploit social
partners (1, 5). Yet, despite its theoretical
importance, the question of why people choose to punish others is still hotly
contested (6). In this study, we use a battery of economic games to
disentangle the different motives underpinning punishment and explore how these
motives vary across individuals.

In humans, many studies of punishment have been carried out in laboratory
settings using economic games (6--15). In these games, participants are usually given a sum of money
that they can use to invest in collective action or to help others.
Alternatively, participants can `cheat' by keeping the money for themselves or
by exploiting the contributions of others. Punishment is introduced into such
games by giving participants the option to pay a small `fee' to impose a greater
`fine' on their co-players. Several lines of experimental evidence indicate that
people use this punishment option (13), that they enjoy
punishing (16), and that they frequently, though not
always (17), punish cheating or exploitative
co-players (11, 12).

Evidence from these experiments suggests that the threat of costly punishment
plays an important role in promoting human cooperation. People tend to cooperate
more in games where punishment is possible compared to those where it is not
(6--8). The effect that the threat of
punishment has on cooperation is also evident in the higher contributions
typically observed in the Ultimatum Game (where punishment is possible) compared
to the structurally-similar Dictator Game (where it is not) (18). This
typical cooperation-enhancing effect of punishment has also been observed across
societies (8), leading some to suggest that costly punishment has
played a key role in the cultural evolution of cooperation in humans
(19--22).

Nevertheless, it remains unclear whether individuals playing economic games use
punishment as a behaviour-change tool to enforce cooperation or as a means to
achieve other ends. Some have argued that punishment is primarily used to shape
future behaviour, either to deter personal harm (3, 10, 23) or to uphold normative standards of cooperative behaviour
(20, 21, 24--27).
But while the \emph{threat} of punishment can have a cooperation-enhancing effect,
the \emph{enactment} of this punishment does not consistently deter targets from
cheating in the future (6). This calls into question whether
punishment primarily operates as a behaviour-change tool or whether it is used
to achieve other goals.

Beyond behaviour shaping concerns, there are a host of other reasons why people
may want to punish in economic games. Punishers might be motivated by a desire
for retribution rather than deterrence, punishing in proportion to the amount
of harm that was personally caused (28). Punishment might be driven
by concerns about relative payoffs, such as disadvantageous inequity aversion
(i.e., avoiding having less than others) (6, 29) and/or
general egalitarian preferences (i.e., wanting all participants to receive the
same payoffs) (30). Such concerns about relative payoffs may be activated
when participants earn less than cheats in economic games or when there are
income disparities in these settings. People might also use punishment for
competitive purposes, seeking advantageous inequity for themselves (i.e., having
more than others) and/or improving their relative position (6).

Common economic game designs have been unable to tease apart these different
motives for punishment because participants who interact with cheats in these
games experience both losses \emph{and} lower relative payoffs. The typical 1:3
fee-fine ratio of punishment in economic games compounds this issue. With
this setup, people can simultaneously use punishment to reciprocate losses, to
deter others from cheating, and to reduce or reverse disparities in payoffs
between themselves and targets. To add to this complexity, it is evident that
people use punishment in seemingly disparate ways: punishing when no behaviour
change is possible, such as in one-shot games (13, 29, 31, 32), on the very last round of repeated games
(33), or in games where the target never learns about the
punishment (34); punishing those who did not cheat or who
over-contributed to collective action (antisocial punishment)
(17, 35); punishing in scenarios where they were not
personally harmed (third-party punishment) (36); and punishing in
scenarios where disparities in payoffs did not arise from participants'
actions (30, 37, 38).

The general conclusion from this research is that there is no one unifying
function of costly punishment in humans. Instead, punishment should be thought
of as a flexible behavioural tool that serves a variety of functions that are
not mutually exclusive (6). Due to its multipurpose nature, we should
therefore expect variation in punishment strategies in the population, much like
the observed variation in social learning strategies (39). Some
individuals may use punishment as a behaviour shaping tool, for example, while
others may use it to reduce or reverse payoff differentials.

This insight raises several underexplored questions. First, which punishment
strategies are more frequent in human populations? Second, what traits predict
adherence to a particular strategy? Previous work has reported that personality
is related to cooperative behaviour (40) and demographics, political
ideology, and religiosity are related to punitive behaviour (41), but
no research has related these variables to specific punishment strategies.
Third, do people have insight into their own punishment strategy? Previous work
has argued that people are often unaware of the underlying function of their
punitive behaviour, yet they feel compelled to enact it anyway
(28, 42).

Here, we aim to delineate nine possible punishment strategies by asking
whether people punish in a manner consistent with a specific strategy and, if
so, what other characteristics (personality, social preferences, political
orientation, religious views) predict the use of different punishment strategies.
Table \ref{tab:tableStrategies} summarises the potential functions for costly
punishment in the economic games that we considered, and the behavioural
strategies they predict. Note that Table \ref{tab:tableStrategies} is not
an exhaustive list of all possible punishment strategies: we do not include
reputational functions of punishment in this table, such as signalling
trustworthiness (4, 43--46),
because our focus is on punishment strategies in anonymous economic games without
reputational incentives {[}but see (47){]}.

Building on previous designs (29, 31, 48, 49), we employ a suite of one-shot economic games where individuals
are given the opportunity to punish targets at a personal cost (Figure
\ref{fig:plotGames}). In each game, targets either steal from another
individual or do nothing. Representative samples of participants from the United
Kingdom (\emph{n} = 1014) and the
United States (\emph{n} = 996)
completed all six games on the online platform Prolific. We carefully designed
the suite of games to tease apart the proposed punishment strategies in Table
\ref{tab:tableStrategies}, such that each strategy predicts a different pattern
of behaviour across all the games (see Methods for more detail about the six
games). We use the resulting behavioural patterns to discern which punishment
strategy participants are employing. We then combine these behavioural patterns
with data on demographics, personality, social preferences, political ideology,
religiosity, and self-reported strategy usage.

@@

Using a suite of economic games measuring punishment in different situations,
we have shown that punishment does not serve just one function, but instead is
a flexible tool that can be and is used for different purposes (6).
While some use punishment to enforce norms of cooperation, others use it to
reduce or even create inequality between individuals. We found that people's
punishment strategy can, to some extent, be predicted by individual differences
in personality, social preferences, and political and religious views. Moreover,
contrary to the view that people are often unable to articulate the reasons for
their punitive behaviour (28), people seem to have some degree of
insight into the strategy they are using. Despite small differences, these
general patterns replicated in samples from both the United Kingdom and the
United States, providing further confidence in the results.

Among the punitive strategies, the most common were particularly sensitive to
inequality in payoffs, either from a self-referential perspective (i.e., avoid
disadvantageous inequity) or more generally (i.e., egalitarian). This is in
line with previous studies which have highlighted inequity aversion as an
important motivation for punishment in economic games
(29--31, 37, 48, 50).
Personality, social preferences, and political and religious views were related
to these strategies. Traits associated with other-regarding concern, such as SVO
and honesty-humility, predicted following the egalitarian strategy. By contrast,
religious and conservative individuals, including individuals higher in SDO and
RWA, were less likely to follow the egalitarian strategy and more likely to
follow the avoid disadvantageous inequity strategy, especially in the United
States. Authoritarian participants in the United States were also more likely to
actively seek advantageous inequity with their punishment. These findings
provide insight into the motives that may have driven higher levels of peer
punishment among conservatives in previous work (41).

Behaviour shaping strategies, such as deterrence and norm-enforcement, were less
common than strategies sensitive to inequality in our set of games. This was
reflected both in participants' elicited punishment behaviour (Figure
\ref{fig:plotModel1b}) and in their self-reports of their own strategy
(Supplementary Figures \ref{fig:plotSliders1} and \ref{fig:plotSliders2}).
Although our design did not explicitly allow for behaviour shaping as the
interactions were all one-shot, we did manipulate whether the target's stealing
behaviour was intentional or not (Game C), an approach which has been used in
previous vignette studies to identify behaviour shaping motives (28).
The lower prevalence of behaviour shaping strategies in our study is consistent
with prior work showing that punishment often continues to be used when
behaviour shaping is impossible, such as when the target will never find out
that they have been punished (34) or on the last round of repeated
interactions (33). We found that demographic, personality, political,
and religious variables tended to be unrelated to behaviour shaping strategies,
though this could reflect low power to detect such associations given the low
prevalence of these strategies in our sample. We also found that participants
accurately reported using the norm-enforcing strategy, but not the deterrent
strategy. This finding is in line with previous research showing that people
struggle to accurately report the deterrent motivations for their punitive
behaviour (28).

We defined antisocial punishment as occurring when a participant exclusively
punished those who did not steal. There were no participants in our sample who
followed this strategy, although some players did punish non-stealing
co-players. Harming non-stealing individuals was also consistent with the
competitive strategies, which did appear in our sample albeit at low frequencies.
The fact that no participants in our sample punished non-stealing across all
games suggests that the antisocial punishment that has been observed in other
studies (17, 51) is not aimed at harming cooperators
specifically, as has been previously suggested (17). Instead,
antisocial punishment is more likely to be motivated by improving one's
relative position, which is in line with work showing that antisocial punishment
disappears when relative payoffs cannot be changed (e.g., when punishment is
only available with a 1:1 fee-fine ratio) (35).

The fact that people use punishment for many different reasons poses problems
for the way that punishment is operationalised in classic behavioural economic
game studies. In these studies, a common assumption is that participants punish
to change the behaviour of cheats (11, 12). But in reality, people
may be choosing the punishment option to achieve a variety of different goals.
This has implications for how people respond to being targeted by punishers in
these games. Targets of punishment in these studies may know that punishment
reflects different motives and can respond accordingly. For example, if targets
interpret punishment as serving a competitive motive, it may elicit retaliation
rather than encourage cooperation (6, 10, 52).
As punishers' motives must be inferred (and such inferences likely depend on
character traits of the target, as well as the context in which punishment
occurs), there is likely to be some variation and error in attributing motives
to punishers. To the extent that inferred motives affect target responses, this
might help to explain the mixed findings in the field as to whether punishment
actually motivates cheating targets to cooperate in the future (6).
